---
title: "multinomialTS — coding demo"
author:
  name: "Quinn Asena, Jack Williams, and Tony Ives"
date: today
bibliography: refs.bib
from: markdown+emoji
format:
  html:
    code-fold: false
    toc: true
    link-external-newwindow: true
    embed-resources: true
execute:
  cache: false
  message: false
theme:
  light: flatly
  dark: darkly
---

### Install packages {.unlisted .hidden}

```{r}
#| label: setup
#| message: false
source("./install_workshop.R")
```

# Introduction

`multinomialTS` fits a state-space model to multinomially distributed ecological
time series (e.g., pollen counts). One of the primary goals is to test multiple
hypotheses about the data: which environmental drivers have the strongest effects
on which taxa, and are interactions among taxa or abiotic variables driving
change in the system? The approach estimates coefficients for taxa interactions
and driver–taxa relationships that we don't get from methods such as ordination
or cluster analysis, and we recommend it as complementary to those methods.

This is a short coding demo. A more detailed walkthrough is in
`state-space-walkthrough.qmd`, and further resources are in the
[main workshop repository](https://github.com/QuinnAsena/multinomialTS-workshop).

**In this demo we will:**

- Read in the pre-wrangled data (taxa already aggregated into focal taxa and groups)
- Plot the state variables and the environmental driver
- Choose a temporal resolution and bin the data
- Find initial conditions with `mnGLMM()`
- Fit `mnTS()` with and without species interactions
- Compare the models

# The data

We work with the built-in `story` data (Story Lake). The taxa have already been
wrangled into focal taxa (*Ulmus*, *Fagus grandifolia*, *Quercus*) plus a
`hardwood` group and an `other` reference group. Full details of that wrangling
step are in the extended walkthrough.

```{r}
#| label: lst-read-data
# State variables (Y): pollen counts, wide format
data(story_pollen_wide, package = "multinomialTS")
# Covariate (X): charcoal accumulation
data(story_char_wide, package = "multinomialTS")
```

The state variables ($Y$):

```{r}
head(story_pollen_wide)
```

And the covariate ($X$):

```{r}
head(story_char_wide)
```

The `age` and `depth` columns are kept for now as common variables to match the
two tibbles; they are removed before fitting.

# State variables $Y$ and the driver

We start by choosing a resolution for the model predictions (e.g., predicting
ecological process at 50- or 100-year intervals).

```{r}
#| label: lst-spacing-checks
# Time-span divided by number of observations
max(story_pollen_wide$age) / nrow(story_pollen_wide)
# The range and average of the age differences between successive observations
range(diff(story_pollen_wide$age))
mean(diff(story_pollen_wide$age))
```

Binning at around 50–75 years keeps most of our data in independent bins. We
apply the same bin-width to both the state variables and the covariate.

## Joining and binning

Charcoal was sampled to a greater depth than the pollen, so we clip the covariate
to the extent of the state variables and join by the common `age` variable.

```{r}
#| label: lst-join
#| warning: false
story_join <- story_char_wide |>
  filter(age <= max(story_pollen_wide$age)) |>
  left_join(story_pollen_wide)

# Always double-check dimensions before/after joining!
dim(story_join)
```

Now we bin. For covariates we take the *average* value per bin; for pollen counts
we take the *sum* (required by the underlying multinomial distribution).

```{r}
#| label: lst-binned
bin_width <- 50
bins <- cut(story_join$age,
            breaks = seq(from = min(story_join$age),
                         to = max(story_join$age + bin_width),
                         by = bin_width), include.lowest = TRUE, labels = FALSE)

story_binned <- bind_cols(bins = bins, story_join) |>
  group_by(bins) |>
  summarise(
    age = mean(age, na.rm = TRUE),        # centre of the bin
    char_acc = mean(char_acc, na.rm = TRUE), # mean charcoal accumulation per bin
    other = sum(other, na.rm = TRUE),     # counts are summed
    hardwood = sum(hardwood, na.rm = TRUE),
    `Fagus grandifolia` = sum(`Fagus grandifolia`, na.rm = TRUE),
    Ulmus = sum(Ulmus, na.rm = TRUE),
    Quercus = sum(Quercus, na.rm = TRUE)
  ) |>
  arrange(desc(age))
# Note: gaps in the pollen data are now filled with 0's, not NA
```

## Plotting the aggregated taxa and the driver

Here are the aggregated state variables (`other`, `hardwood`, and the three focal
taxa) alongside the environmental driver (`char_acc`), at the `r bin_width`-year
resolution that goes into the model. Note the different scale for charcoal
accumulation versus pollen counts.

```{r}
#| label: fig-taxa-driver
#| code-fold: true
#| warning: false
#| fig-cap: "Binned state variables (pollen counts of the focal taxa and the 'hardwood' and 'other' groups) and the environmental driver (charcoal accumulation), at 50-year resolution."
#| fig-width: 9
#| fig-height: 5

demo_long <- story_binned |>
  select(age, char_acc, other, hardwood, Ulmus, `Fagus grandifolia`, Quercus) |>
  pivot_longer(-age, names_to = "variablename", values_to = "value") |>
  mutate(variablename = factor(
    variablename,
    levels = c("char_acc", "other", "hardwood", "Ulmus", "Fagus grandifolia", "Quercus"),
    labels = c("Charcoal acc.", "other", "hardwood", "Ulmus", "Fagus", "Quercus")
  ))

ggplot(demo_long, aes(x = age, y = value)) +
  geom_area(colour = "grey90") +
  geom_segment(aes(x = age, xend = age, y = 0, yend = value),
               colour = "grey30", linewidth = 0.6) +
  scale_x_reverse(breaks = scales::breaks_pretty(n = 6)) +
  coord_flip() +
  labs(y = "Pollen counts / charcoal accumulation", x = "Time (ybp)") +
  facet_wrap(~variablename, nrow = 1, scales = "free_x") +
  theme_minimal() +
  theme(text = element_text(size = 10))
```

## Building the model matrices

`multinomialTS` requires two matrices: a site-by-species matrix $Y$, and a
covariate matrix $X$. $X$ variables may be of different types but should be scaled
relative to each other.

```{r}
#| label: lst-matrices
story_pollen_matrix <- story_binned |>
  select(other, hardwood, Ulmus, `Fagus grandifolia`, Quercus) |>
  rename("Fagus" = "Fagus grandifolia") |>
  as.matrix()
# Empty (gap) rows are set to NA so they are not treated as zero-count observations
story_pollen_matrix[which(rowSums(story_pollen_matrix) == 0), ] <- NA
head(story_pollen_matrix)

story_char_matrix_scaled <- story_binned |>
  select(char_acc) |>
  as.matrix() |>
  scale()
head(story_char_matrix_scaled)
```

# Fitting `multinomialTS`

We now have a temporal resolution of `r bin_width`-year bins, a site-by-species
matrix $Y$, and a binned and scaled covariate matrix $X$.

## Finding initial conditions with `mnGLMM()`

`mnTS()` estimates biotic interactions ($C$ matrix), taxa-driver relationships
($B$ matrix), and cross-correlated error ($V$ matrix), but it needs starting
values. We get those from `mnGLMM()`, which returns estimates for $B$ and $V$
(not $C$) and assumes no time gaps in the data.

::: {.callout-tip}
Arguments for *starting* values are suffixed `.start`; arguments for the
parameters to be *estimated* are suffixed `.fixed`.
:::

```{r}
#| label: lst-sample-idx
# Row indices of bins that contain pollen observations (exclude empty/gap bins)
sample_idx <- which(rowSums(story_pollen_matrix) != 0)
```

```{r}
#| label: lst-glmm-params
p <- ncol(story_char_matrix_scaled) + 1 # number of covariates + intercept
n <- ncol(story_pollen_matrix)          # number of taxa
V.fixed.glmm <- diag(n)
diag(V.fixed.glmm) <- NA
V.fixed.glmm[1] <- 1
B.fixed.glmm <- matrix(c(rep(0, p), rep(NA, (n - 1) * p)), p, n) # reference taxa [,1] set to 0
B.start.glmm <- matrix(c(rep(0, p), rep(.01, (n - 1) * p)), p, n)
```

`mnGLMM()` does not handle gaps, so we index the matrices by `sample_idx`. The fit
below is precomputed (loaded from `./outputs/`) so the document renders quickly;
set `eval: true` to run it live.

```{r}
#| label: lst-fit-glmm
#| warning: false
#| eval: false
start_time <- Sys.time()
glmm_mod <- mnGLMM(Y = story_pollen_matrix[sample_idx, ],
                   X = story_char_matrix_scaled[sample_idx, , drop = FALSE],
                   B.start = B.start.glmm,
                   B.fixed = B.fixed.glmm,
                   V.fixed = V.fixed.glmm)
end_time <- Sys.time()
end_time - start_time
saveRDS(glmm_mod, "./outputs/glmm_mod.rds")
```

```{r}
glmm_mod <- readRDS("./outputs/glmm_mod.rds")
summary(glmm_mod)
```

## Setting up parameters for `mnTS()`

We use the `mnGLMM()` output as starting values for $B0$, $B$, and $V$. `mnGLMM()`
does not estimate $C$, so we seed $C$ with values close to zero and let `mnTS()`
do the rest.

```{r}
#| label: lst-mnts-params
B0.start.mnTS <- glmm_mod$B[1, , drop = FALSE] # intercept
B.start.mnTS  <- glmm_mod$B[2, , drop = FALSE] # covariate; add rows for more covariates

sigma.start.mnTS <- glmm_mod$sigma

V.fixed.mnTS <- matrix(NA, n, n)
V.fixed.mnTS[1] <- 1
V.start.mnTS <- glmm_mod$V

B.fixed.mnTS <- matrix(NA, p - 1, n)
B.fixed.mnTS[, 1] <- 0
B0.fixed.mnTS <- matrix(c(0, rep(NA, n - 1)), nrow = 1, ncol = n)

# C without interactions: diagonal = self-regulation (estimated); off-diagonal = 0 (not estimated)
C.start.mnTS <- .5 * diag(n)
C.fixed.mnTS <- C.start.mnTS
C.fixed.mnTS[C.fixed.mnTS != 0] <- NA

# C with an interaction between Fagus and Quercus
C.start.int.mnTS <- .5 * diag(n)
C.start.int.mnTS[5, 4] <- .001 # effect of Fagus (col 4) on Quercus (row 5)
C.start.int.mnTS[4, 5] <- .001 # effect of Quercus (col 5) on Fagus (row 4)
C.fixed.int.mnTS <- C.start.int.mnTS
C.fixed.int.mnTS[C.fixed.int.mnTS != 0] <- NA
```

## Fitting `mnTS()`

Unlike `mnGLMM()`, `mnTS()` *does* handle gaps in the state variables where there
are data in the covariate matrix. We fit the model twice: without and with taxa
interactions. Both fits are precomputed; set `eval: true` to run them live.

::: {.panel-tabset}

## Without interactions

```{r}
#| label: lst-fit-mnts
#| eval: false
mnTS_mod <- mnTS(Y = story_pollen_matrix[sample_idx, ],
                 X = story_char_matrix_scaled, Tsample = sample_idx,
                 B0.start = B0.start.mnTS, B0.fixed = B0.fixed.mnTS,
                 B.start = B.start.mnTS, B.fixed = B.fixed.mnTS,
                 C.start = C.start.mnTS, C.fixed = C.fixed.mnTS,
                 V.start = V.start.mnTS, V.fixed = V.fixed.mnTS,
                 dispersion.fixed = 1, maxit.optim = 1e+6)
# maxit.optim: max optimiser iterations before stopping; increase if the model needs more time
saveRDS(mnTS_mod, "./outputs/mnTS_mod.rds")
```

```{r}
mnTS_mod <- readRDS("./outputs/mnTS_mod.rds")
```

## With interactions

```{r}
#| label: lst-fit-mnts-int
#| eval: false
mnTS_mod_int <- mnTS(Y = story_pollen_matrix[sample_idx, ],
                     X = story_char_matrix_scaled, Tsample = sample_idx,
                     B0.start = mnTS_mod$B0, B0.fixed = B0.fixed.mnTS,
                     B.start = mnTS_mod$B, B.fixed = B.fixed.mnTS,
                     C.start = mnTS_mod$C, C.fixed = C.fixed.int.mnTS,
                     V.start = mnTS_mod$V, V.fixed = V.fixed.mnTS,
                     dispersion.fixed = 1, maxit.optim = 1e+6)
saveRDS(mnTS_mod_int, "./outputs/mnTS_mod_int.rds")
```

```{r}
mnTS_mod_int <- readRDS("./outputs/mnTS_mod_int.rds")
```

:::

# Interpreting the output

`coef()` and `summary()` show the model output. Here are the coefficients of the
interaction model:

```{r}
coef(mnTS_mod_int)
```

We have found that bootstrapping provides better estimates of standard errors
(and P-values). The `boot()` function bootstraps the model but can take a long
time; we don't do it here, but we strongly recommend bootstrapping your final
models — see the [full walkthrough](state-space-walkthrough.qmd).

## Comparing models

A quick comparison of AIC between the model without and with interactions:

```{r}
data.frame(without_interactions = mnTS_mod$AIC,
           with_interactions = mnTS_mod_int$AIC)
```

Which model gives the lowest AIC? Does that agree with what you might expect from
the data, and what hypothesis are we testing by adding the interaction?
