---
title: "multinomialTS coding demo"
author:
  name: "Quinn Asena • Jack Williams • Tony Ives"
date: today
bibliography: refs.bib
from: markdown+emoji
format:
  html:
    code-fold: false
    toc: true
    link-external-newwindow: true
    embed-resources: true
execute:
  cache: false
  message: false
theme:
  light: flatly
  dark: darkly
---

### Install packages {.unlisted .hidden}

```{r}
#| label: setup
#| message: false
source("./install_workshop.R")
```

# Introduction

`multinomialTS` fits a state-space model to multinomially distributed ecological
time series (community count data such as pollen counts). One of the primary goals is to test multiple
hypotheses about the data: which environmental drivers have the strongest effects
on which taxa, and are interactions among taxa or abiotic variables driving
change in the system? The approach estimates coefficients for taxa interactions
and driver–taxa relationships that we don't get from methods such as ordination
or cluster analysis, and we recommend it as complementary to those methods.

This is a short coding demo. A more detailed walkthrough is in
`state-space-walkthrough.qmd`, and further resources are in the
[main workshop repository](https://github.com/QuinnAsena/multinomialTS-workshop).
An HTML version of the full walkthrough is hosted at
[multinomialTS-workshop/](https://quinnasena.github.io/multinomialTS-workshop/state-space-walkthrough.html).
In this demo we will use the example data in the package to:

- Choose a temporal resolution for prediction
- Find initial conditions with `mnGLMM()`
- Fit `mnTS()` with and without species interactions

# The data

We will work with the built-in `story` data (Story Lake, IN; [@schlenker2024]). The taxa have already been
wrangled into focal taxa (*Ulmus* / elm, *Fagus grandifolia* / American beech, *Quercus* / oak) plus a
`hardwood` group and an `other` reference group. Full details of that wrangling
step are in the extended walkthrough.

```{r}
#| label: lst-read-data
# State variables (Y): pollen counts, wide format
data(story_pollen_wide, package = "multinomialTS")
# Covariate (X): charcoal accumulation
data(story_char_wide, package = "multinomialTS")
```

The state variables ($Y$):

```{r}
head(story_pollen_wide)
```

And the covariate ($X$):

```{r}
head(story_char_wide)
```

I keep the `age` and `depth` columns for now as common variables to join the
two tibbles later; they are removed before fitting.

## Plotting the raw data

First, let's plot the pollen count data. Charcoal is sampled at a finer temporal resolution than the pollen data; this is ideal for the model, which can handle a finer temporal resolution of covariates.

```{r}
#| label: fig-raw-data
#| code-fold: true
#| warning: false
#| fig-cap: "Raw (unbinned) data: pollen counts of the focal taxa and the 'hardwood' and 'other' groups (left), and the charcoal accumulation rate covariate (right)."
#| fig-width: 10
#| fig-height: 8

# State variables: pivot the aggregated taxa to long format
# (ungroup first: story_pollen_wide is grouped by depth, which select() would otherwise re-add)
species_long <- story_pollen_wide |>
  ungroup() |>
  select(age, other, hardwood, Ulmus, `Fagus grandifolia`, Quercus) |>
  pivot_longer(-age, names_to = "variablename", values_to = "value") |>
  mutate(variablename = factor(
    variablename,
    levels = c("other", "hardwood", "Ulmus", "Fagus grandifolia", "Quercus"),
    labels = c("other", "hardwood", "Ulmus", "Fagus", "Quercus")
  ))

p_species <- ggplot(species_long, aes(x = age, y = value)) +
  geom_area(colour = "grey90") +
  geom_segment(aes(x = age, xend = age, y = 0, yend = value),
               colour = "grey30", linewidth = 0.6) +
  scale_x_reverse(breaks = scales::breaks_pretty(n = 6)) +
  coord_flip() +
  labs(y = "Pollen counts", x = "Time (ybp)") +
  facet_wrap(~variablename, nrow = 1) +
  theme_minimal() +
  theme(text = element_text(size = 10))

# Covariate: clip charcoal to the pollen extent so the age axes align
p_char <- story_char_wide |>
  filter(age <= max(story_pollen_wide$age)) |>
  ggplot(aes(x = age, y = char_acc)) +
  geom_line(colour = "grey40", linewidth = .6) +
  geom_point(size = 1) +
  scale_x_reverse(breaks = scales::breaks_pretty(n = 6)) +
  coord_flip() +
  labs(y = "Charcoal acc. rate", x = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 10),
        axis.text.y = element_blank())

# Combine the species panels and the covariate panel with patchwork
p_species + p_char + plot_layout(widths = c(.8, .2))
```

# State variables $Y$ and the driver

We start by choosing a resolution for the model predictions (e.g., predicting
ecological process at 50- or 100-year intervals).

```{r}
#| label: lst-spacing-checks
# Time-span divided by number of observations
max_over_n <- max(story_pollen_wide$age) / nrow(story_pollen_wide)
# The range and average of the age differences between successive observations
diff_rng <- range(diff(story_pollen_wide$age))
diff_mean <- mean(diff(story_pollen_wide$age))

cat(
  "\n",
  "Max age / nrow = ", max_over_n, "\n",
  "Range of age differences = ", diff_rng[1], " to ", diff_rng[2], "\n",
  "Mean age difference = ", diff_mean, "\n"
)
```

Binning at around 50–75 years keeps most of our data in independent bins. We
apply the same bin-width to both the state variables and the covariate.

## Joining and binning

Charcoal was sampled at a finer resolution, and to a greater depth than the pollen, so we clip the covariate
to the extent of the state variables and join by the common `age` and `depth` variables.

```{r}
#| label: lst-join
#| warning: false
story_join <- story_char_wide |>
  filter(age <= max(story_pollen_wide$age)) |>
  left_join(story_pollen_wide)

# Always double-check dimensions before/after joining!
dim(story_join)
# Note that there are now gaps in the pollen data but not the charcoal
head(story_join)
```

::: {.callout-tip}
## Ask me questions!
What questions do you have so far?
:::

Once we have chosen a prediction resolution, we can bin the data. 
For covariates we take the *average* value per bin; for pollen counts
we take the *sum* (required by the underlying multinomial distribution).

::: {.callout-tip}
## Note

We can have a finer temporal resolution of the covariates, but we _cannot_ have a finer resolution of the state variables. If your data have a finer resolution of the state variables then it is possible to interpolate the covariate data to match.
:::

```{r}
#| label: lst-binned
bin_width <- 50
bins <- cut(story_join$age,
            breaks = seq(from = min(story_join$age),
                         to = max(story_join$age + bin_width),
                         by = bin_width), include.lowest = TRUE, labels = FALSE)

story_binned <- bind_cols(bins = bins, story_join) |>
  group_by(bins) |>
  summarise(
    age = mean(age, na.rm = TRUE),        # centre of the bin
    char_acc = mean(char_acc, na.rm = TRUE), # mean charcoal accumulation per bin
    other = sum(other, na.rm = TRUE),     # counts are summed
    hardwood = sum(hardwood, na.rm = TRUE),
    `Fagus grandifolia` = sum(`Fagus grandifolia`, na.rm = TRUE),
    Ulmus = sum(Ulmus, na.rm = TRUE),
    Quercus = sum(Quercus, na.rm = TRUE)
  ) |>
  arrange(desc(age))
# Note: gaps in the pollen data are now filled with 0's, not NA
```

::: {.callout-important}
## Important
We want the model to run from the past to the present, and the model runs from the top of the matrix to the bottom. So we want to make sure that the top row of the data is the _oldest_ date. Palaeo-data are usually ordered from the top of the core to the bottom (present to past), so we `arrange(desc(age))`.
:::



## Building the model matrices

`multinomialTS` requires two matrices: a site-by-species matrix $Y$, and a
covariate matrix $X$. $X$ variables may be of different types but should be scaled
relative to each other.

```{r}
#| label: lst-matrices
story_pollen_matrix <- story_binned |>
  select(other, hardwood, Ulmus, `Fagus grandifolia`, Quercus) |>
  rename("Fagus" = "Fagus grandifolia") |>
  as.matrix()
# Empty (gap) rows are set to NA so they are not treated as zero-count observations
story_pollen_matrix[which(rowSums(story_pollen_matrix) == 0), ] <- NA
head(story_pollen_matrix)

story_char_matrix_scaled <- story_binned |>
  select(char_acc) |>
  as.matrix() |>
  scale()
head(story_char_matrix_scaled)
```

# Fitting `multinomialTS`

We now have a temporal resolution of `r bin_width`-year bins, a site-by-species
matrix $Y$, and a binned and scaled covariate matrix $X$.

## Finding initial conditions with `mnGLMM()`

`mnTS()` estimates biotic interactions ($C$ matrix), taxa-driver relationships
($B$ matrix), and cross-correlated error ($V$ matrix), but it needs starting
values for $B$, $B0$ (intercept), and $V$. We get those from `mnGLMM()`, which
returns estimates for $B$ and $V$ (not $C$) and assumes no time gaps in the data.

For `mnGLMM()` we need:

- A vector that indicates the row indices where there are state variable observations: `sample_idx`
- An integer number of covariates (+ 1 for `mnGLMM()`): `p` 
- An integer of the number of state variables: `n`
- A matrix of starting values for $B$: `B.start.glmm`
- A matrix of $B$ parameters to estimate: `B.fixed.glmm` 
- A matrix of $V$ parameters to estimate: `V.fixed.glmm`

::: {.callout-tip}
Arguments for *starting* values are suffixed `.start`; arguments for the
parameters to be *estimated* are suffixed `.fixed`.
:::

```{r}
#| label: lst-sample-idx
# Row indices of bins that contain pollen observations (exclude empty/gap bins)
sample_idx <- which(rowSums(story_pollen_matrix) != 0)
```

```{r}
#| label: lst-glmm-params
p <- ncol(story_char_matrix_scaled) + 1 # number of covariates + intercept
n <- ncol(story_pollen_matrix)          # number of taxa
V.fixed.glmm <- diag(n)
diag(V.fixed.glmm) <- NA
V.fixed.glmm[1] <- 1
B.fixed.glmm <- matrix(c(rep(0, p), rep(NA, (n - 1) * p)), p, n) # reference taxa [,1] set to 0
B.start.glmm <- matrix(c(rep(0, p), rep(.01, (n - 1) * p)), p, n)
```

`mnGLMM()` does not handle gaps, so we index the matrices by `sample_idx`. The fit
below is precomputed (loaded from `./outputs/`) so the document renders quickly;
set `eval: true` to run it live.

```{r}
#| label: lst-fit-glmm
#| warning: false
#| eval: false
start_time <- Sys.time()
glmm_mod <- mnGLMM(Y = story_pollen_matrix[sample_idx, ],
                   X = story_char_matrix_scaled[sample_idx, , drop = FALSE],
                   B.start = B.start.glmm,
                   B.fixed = B.fixed.glmm,
                   V.fixed = V.fixed.glmm)
end_time <- Sys.time()
end_time - start_time
saveRDS(glmm_mod, "./outputs/glmm_mod.rds")
```

```{r}
glmm_mod <- readRDS("./outputs/glmm_mod.rds")
summary(glmm_mod)
```

## Setting up parameters for `mnTS()`

We use the `mnGLMM()` output as starting values for $B0$, $B$, and $V$. `mnGLMM()`
does not estimate $C$, so we seed $C$ with values close to zero and let `mnTS()`
do the rest.

```{r}
#| label: lst-mnts-params
B0.start.mnTS <- glmm_mod$B[1, , drop = FALSE] # intercept
B.start.mnTS  <- glmm_mod$B[2, , drop = FALSE] # covariate; add rows for more covariates

sigma.start.mnTS <- glmm_mod$sigma

V.fixed.mnTS <- matrix(NA, n, n)
V.fixed.mnTS[1] <- 1
V.start.mnTS <- glmm_mod$V

B.fixed.mnTS <- matrix(NA, p - 1, n)
B.fixed.mnTS[, 1] <- 0
B0.fixed.mnTS <- matrix(c(0, rep(NA, n - 1)), nrow = 1, ncol = n)

# C without interactions: diagonal = self-regulation (estimated); off-diagonal = 0 (not estimated)
C.start.mnTS <- .5 * diag(n)
C.fixed.mnTS <- C.start.mnTS
C.fixed.mnTS[C.fixed.mnTS != 0] <- NA

# C with an interaction between Fagus and Quercus
C.start.int.mnTS <- .5 * diag(n)
C.start.int.mnTS[5, 4] <- .001 # effect of Fagus (col 4) on Quercus (row 5)
C.start.int.mnTS[4, 5] <- .001 # effect of Quercus (col 5) on Fagus (row 4)
C.fixed.int.mnTS <- C.start.int.mnTS
C.fixed.int.mnTS[C.fixed.int.mnTS != 0] <- NA
```

## Fitting `mnTS()`

Unlike `mnGLMM()`, `mnTS()` *does* handle gaps in the state variables where there
are data in the covariate matrix. We fit the model twice: without and with taxa
interactions. Both fits are precomputed; set `eval: true` to run them live.

::: {.panel-tabset}

## Without interactions

```{r}
#| label: lst-fit-mnts
#| eval: false
mnTS_mod <- mnTS(Y = story_pollen_matrix[sample_idx, ],
                 X = story_char_matrix_scaled, Tsample = sample_idx,
                 B0.start = B0.start.mnTS, B0.fixed = B0.fixed.mnTS,
                 B.start = B.start.mnTS, B.fixed = B.fixed.mnTS,
                 C.start = C.start.mnTS, C.fixed = C.fixed.mnTS,
                 V.start = V.start.mnTS, V.fixed = V.fixed.mnTS,
                 dispersion.fixed = 1, maxit.optim = 1e+6)
# maxit.optim: max optimiser iterations before stopping; increase if the model needs more time
saveRDS(mnTS_mod, "./outputs/mnTS_mod.rds")
```

```{r}
mnTS_mod <- readRDS("./outputs/mnTS_mod.rds")
```

## With interactions

```{r}
#| label: lst-fit-mnts-int
#| eval: false
mnTS_mod_int <- mnTS(Y = story_pollen_matrix[sample_idx, ],
                     X = story_char_matrix_scaled, Tsample = sample_idx,
                     B0.start = mnTS_mod$B0, B0.fixed = B0.fixed.mnTS,
                     B.start = mnTS_mod$B, B.fixed = B.fixed.mnTS,
                     C.start = mnTS_mod$C, C.fixed = C.fixed.int.mnTS,
                     V.start = mnTS_mod$V, V.fixed = V.fixed.mnTS,
                     dispersion.fixed = 1, maxit.optim = 1e+6)
saveRDS(mnTS_mod_int, "./outputs/mnTS_mod_int.rds")
```

```{r}
mnTS_mod_int <- readRDS("./outputs/mnTS_mod_int.rds")
```

:::

# Interpreting the output

`coef()` and `summary()` show the model output:

::: {.panel-tabset}

## Without interactions

```{r}
coef(mnTS_mod)
```


## With interactions
```{r}
coef(mnTS_mod_int)
```

:::

We have found that bootstrapping provides better estimates of standard errors
(and P-values). The `boot()` function bootstraps the model but can take a long
time; we don't do it here, but we strongly recommend bootstrapping your final
models — see the [full walkthrough](state-space-walkthrough.qmd).


# References

::: {#refs}
:::